Nonparametric Clustering from Energy Statistics

### 2017-07-18 qcqp_clustering/draft/energy_clustering.tex
still working on draft

\usepackage{algorithmic}
\usepackage{algorithm}

@@

\begin{abstract}
Energy statistics provides a nonparametric test for equality of distributions.
It was proposed by 
Sz\' ekely in the 80's
inspired by the Newtonian gravitational potential from classical mechanics. 
The idea
is to associate a statistical potential energy to observations, such that 
minimum energy is achieved under the null hypothesis of equality of 
distributions. Energy statistics
was further generalized to probability 
distributions on arbitrary metric spaces,
and more recently, a connection with kernels in RKHS was established.
Nevertheless, although extensively used by the statistics community, energy
statistics has not been
previously incorporated in machine learning problems.
In this paper, we consider the problem of clustering data from
an energy statistics theory perspective.
We provide a precise mathematical formulation, yielding
a quadratically constrained quadratic optimization problem (QCQP). 
We show that
this is equivalent to kernel $k$-means
optimization problem, however,
energy statistics is able to fix the kernel choice. 
This clustering 
method is nonparametric, and if prior information is available
it can be easily incorporated in the kernel construction.
We propose an iterative algorithm
to find local optimizers of this QCQP problem, based on the change
in the statistical energy 
of moving points to different clusters. This algorithm 
is different but has the same computational cost as kernel $k$-means 
algorithm.
We then compare this algorithm with
well-known kernel $k$-means, standard $k$-means, and GMM algorithms by 
providing carefully designed numerical experiments. The results
show that, in general, this method outperforms
these most used clustering algorithms.
\end{abstract}

@@

Energy statistics is based on the energy distance between probability
distributions, which provides a notion of statistical potential energy
to statistical observations, in close analogy to Newton's gravitational
potential energy in classical mechanics. We refer the reader to
\cite{Szkely2013}, and references therein, for an overview.
It has been applied to several goodness-of-fit hypothesis tests, 
multi-sample tests of equality of distributions, analysis of variance
\cite{RizzoVariance}, and (nonlinear)
dependence tests through
distance covariance and distance correlation, which generalizes the Pearson
correlation coefficient. 
Moreover, energy statistics  was applied
to hierarchical clustering \cite{RizzoClustering} by extending Ward's
method of minimum variance.

More recently, the distance covariance was 
generalized from Euclidean
spaces to metric spaces of strong negative type \cite{Lyons}.
Furthermore, a unifying framework establishing an equivalence
between generalized energy distances to maximum
mean discrepancies (MMD), which are distances between embeddings of
distributions in reproducing kernel Hilbert spaces (RKHS), was
provided \cite{Sejdinovic2013}. This important work establishes the link
between techniques commonly used in the statistics literature, regarding
energy statistics, and techniques
commonly used in machine learning.

Given a dataset $\mathbb{X} = \{x_1, x_2,\dotsc,x_n \}$, where each
datapoint lives in a generic space $x_i \in \mathcal{X}$, the $k$-clustering
problem consists in grouping these points into $k$ groups $\{
\C_1,\C_2,\dotsc,\C_k \}$, such that
points belonging to the same group are more ``similar'' to each other
than to points in  other groups. This intuitive notion already assumes
a ``metric'' able to measure the similarity between datapoints. Clustering
is an unsupervised method, and one of the most important problems in machine
learning since it provides the first step towards automatically constructing
labels for previously unseen data. The most used algorithms are
by far $k$-means and gaussian mixture models (GMM) through expectation
maximization (EM) algorithm. Both are parametric, making 
strong assumptions about
the distribution of the data (normality), and the 
space where data lie is Euclidean, $\mathcal{X} = \mathbb{R}^D$, hence
the similarity is based on Euclidean metric, 
$\| x_i - x_j \|^2 = (x_i - x_j)^\top(x_i-x_j) = 
\sum_{\ell=1}^D (x_{i,\ell} - x_{j,\ell})^2$. These algorithms provide
a good clustering when data is linearly separable in Euclidean space.

To account for nonlinearly separable data, which may possibly live
in an arbitrary non-Euclidean space, kernel methods are usually employed.
If a Mercer kernel $\kk: \mathcal{X}\times \mathcal{X} \to \mathbb{R}$ 
is available \cite{Mercer}, which is a symmetric and positive definite 
function, it guarantees that there exists a map $\varphi : \mathcal{X} \to
\mathcal{H}$, where $\mathcal{H}$ is a Hilbert space. Thus, one can
compute similarities through the inner product and 
perform clustering in the feature space $\mathcal{H}$. The nice thing
about this is that the map $\varphi$ is only used implicitly, i.e. one
does not need to know $\varphi$, since the inner
product can be computed only based on the kernel $\kk(\cdot, \cdot)$, 
a convenient technique known as kernel trick. The well-known 
kernel $k$-means problem 
is exactly $k$-means in the feature space
\cite{Smola,Girolami} (see also \cite{Filippone} for a survey of clustering
methods). 
It exploits the nonlinear structure provided by
a kernel function to perform clustering in datasets that are not linearly
separable in their original space $\mathcal{X}$.
As it stands, kernel $k$-means is an heuristic approach in the sense that
it is not derived from a statistical theory. Moreover, the choice of
kernel is crucial, and there is no systematic theory for this. One
must rely on ad hoc methods.

In this paper, we consider how to perform clustering based on energy
statistics theory. The original goal is to provide a nonparametric
clustering method, since energy distance is nonparametric.
The basis of our work is the theory developed in 
\cite{Sejdinovic2013}. Based on this and on the multi-sample test
for equality of distributions from energy statistics, we show that
there is only one function that provides the clustering optimization
problem. We then show that this clustering
optimization problem reduces to a quadratically constrained quadratic
problem (QCQP). This optimization problem has the same form as kernel $k$-means,
spectral clustering, and some well-known graph partitioning problems
\cite{Dhillon}. This lead us to show that energy statistics based
clustering is actually equivalent to kernel $k$-means optimization problem.
However, energy statistics fixes a kernel to begin with.
Furthermore, these results can be seen as a first-principle derivation
of kernel $k$-means from energy statistics, which thus bring
these kernel based clustering methods into a unifying mathematical theory.
We also provide an iterative algorithm, different but with the same
complexity as kernel $k$-means algorithm. The numerical results indicate
that this method is more accurate and robust.

Our work is organized as follows. In section~\ref{sec:background} we review
the necessary background on energy statistics and RKHS.
Section~\ref{sec:clustering_theory} contains the main results of this paper
where we consider a clustering theory based on energy statistics, leading
to a QCQP which is NP-hard.
In section~\ref{sec:twoclass} we consider a simple example in one dimension,
where we propose an algorithm which requires no initialization.
In section~\ref{sec:algo} we briefly review kernel $k$-means algorithm,
and propose a new iterative algorithm to solve this QCQP.
Section~\ref{sec:numerics} contains some carefully designed numerical
experiments indicating that this algorithm outperforms kernel
$k$-means, standard $k$-means, and GMM algorithms.
Our conclusions are in section~\ref{sec:conclusion}.

@@

In this section we briefly review the main concepts from energy
statistics and its relation to reproducing kernel Hilbert spaces 
(RKHS) which form the basis of our work.
For more details we refer the reader
to \cite{Szkely2013} (and references therein) and 
also \cite{Sejdinovic2013}.

Consider random variables in $\mathbb{R}^D$ 
such that $X,X' \stackrel{iid}{\sim} P$ and 
$Y,Y' \stackrel{iid}{\sim} Q$, where $P$ and $Q$ are cumulative
distribution functions with finite first moments. 
The quantity \cite{Szkely2013}
\begin{equation}
\label{eq:energy}
\Energy(P, Q) \equiv 2 \E \| X - Y\| - \E \| X - X' \| - \E \| Y - Y' \|,
\end{equation}
called \emph{energy distance}, 
is rotational invariant and nonnegative, $\Energy(P,Q) \ge 0$, where
equality
to zero holds if and only if $P = Q$.
Above $\| \cdot \|$ denotes the
Euclidean norm in $\mathbb{R}^D$. 
Energy distance
provides a characterization of equality of distributions, and
$\Energy^{1/2}$ is
a metric on the space of distributions.

@@

It is important to 
mention that \eqref{eq:energy2} can be even further generalized.
Let $X, Y \in \mathcal{X}$,  where $\mathcal{X}$ is a space endowed
with a \emph{semimetric of negative type}
$\rho: \mathcal{X}\times\mathcal{X} \to \mathbb{R}$, which is required
to satisfy
\begin{equation}
\label{eq:negative_type}
\sum_{i,j=1}^n \alpha_i \alpha_j \rho(X_i, X_j) \le 0,
\end{equation}
where $X_i \in \mathcal{X}$ and $\alpha_i \in \mathbb{R}$ such that
$\sum_{i=1}^n \alpha_i = 0$. Then, $\mathcal{X}$ is called a \emph{space of
negative type}.
We can thus replace $\mathbb{R}^D \to \mathcal{X}$ and 
$\| X - Y \| \to \rho(X , Y)$ in the definition \eqref{eq:energy}, obtaining
the energy distance as
\begin{equation}
\label{eq:energy3}
\Energy(P, Q) \equiv 2 \E \rho(X,Y) - \E \rho(X, X') - \E \rho(Y,Y').
\end{equation}
For spaces of negative type, there exists a Hilbert space $\mathcal{H}$ and
a map $\varphi: \mathcal{X} \to
\mathcal{H}$ such that
$\rho(X, Y) = \| \varphi(X) - \varphi(Y) \|_{\mathcal{H}}^2$, which
allows us to compute quantities related to probability distributions over
$\mathcal{X}$ in the Hilbert space $\mathcal{H}$.
Even though the semimetric 
$\rho$ may not satisfy the triangle inequality, 
$\rho^{1/2}$ does since it can be shown to be a legit metric.

There is an equivalence between energy distance, 
commonly used in statistics,
and distances between embeddings of distributions in 
RKHS, commonly used in machine learning. 
This equivalence was established
in \cite{Sejdinovic2013}. We first recall the definition of
RKHS. Let $\HH$ be a Hilbert space of real-valued functions
over $\mathcal{X}$. A function 
$\kk : \mathcal{X} \times \mathcal{X} \to 
\mathbb{R}$ is a reproducing kernel of $\HH$ if it satisfies
the following two conditions:
\begin{enumerate}
\item $\kkk_x \equiv \kk(\cdot, x) \in \HH$ 
for all $x \in \mathcal{X}$.
\item $\langle \kkk_x, f \rangle_{\HH} = f(x)$ for
all $x\in\mathcal{X}$ and $f\in \HH$.
\end{enumerate}
In other words, for any $x \in \mathcal{X}$ and any function $f \in \HH$,
there is a unique 
$\kkk_x \in \HH$ that reproduces $f(x)$ through the inner product
of $\HH$.
If such a \emph{kernel} 
function $\kk$ exists, then $\HH$ is called a RKHS. The above two 
properties immediately imply that $\kk$ is symmetric and positive
definite. Indeed, notice that
$\langle \kkk_x, \kkk_y \rangle = \kkk_y(x) = \kk(x,y)$, and since
this inner product is real,
$\langle \kkk_x, \kkk_y \rangle^* = \langle \kkk_y, \kkk_x \rangle = 
\langle \kkk_x, \kkk_y \rangle$, we immediately have that
the kernel is symmetric,
$\kk(y,x) = \kk(y,x)$. Moreover, for any $w \in
\HH$ we can write $w = \sum_{i=1}^n c_i \kkk_{x_i}$, where
$\{ \kkk_{x_i} \}_{i=1}^n$ is a basis of $\HH$. It follows that
$\langle w, w \rangle_{\HH}  = \sum_{i,j=1}^n c_i c_j \kk(x_i,x_j) \ge 0$,
showing that the kernel is positive definite. If $G$ is a matrix with
elements $G_{ij} = \kk(x_i,x_j)$, this is equivalent to $G$ being
positive semi-definite: $\bm{v}^\top G \, \bm{v} \ge 0$ for any vector
$\bm{v} \in \mathbb{R}^n$.

The Moore-Aronszajn theorem 
\cite{Aronszajn}
establishes the converse of the above paragraph. 
For every symmetric
and positive definite function $\kk: \mathcal{X}\times \mathcal{X} \to
\mathbb{R}$, there is an associated RKHS $\Hk$ 
with reproducing
kernel $\kk$. The map $\varphi: x \mapsto \kkk_x \in \Hk$ is called
the canonical \emph{feature map}. Given a kernel $\kk$,
this theorem enables us to define an embedding of a probability measure
$P$ into the RKHS: $P \mapsto \kkk_P \in
\Hk$ such that 
$\int f(x) d P(x) = \langle f, \kkk_P \rangle$ for all $f \in \Hk$,
or alternatively, $\kkk_P \equiv \int \kk( \, \cdot \,, x)  d P(x)$. 
We can now  introduce the 
notion of distance between two probability measures using the inner product
of $\Hk$. This is called the maximum mean discrepancy (MMD) and
is given by
\begin{equation}\label{eq:mmd}
\gamma_\kk(P,Q) \equiv \| \kkk_P - \kkk_Q \|_{\Hk},
\end{equation}
which can also be written as \cite{Gretton2012}
\begin{equation}\label{eq:mmd2}
\gamma_\kk^2(P,Q) = \E \kk(X,X') + \E \kk(Y,Y') - 2 \E \kk(X, Y)
\end{equation}
where $X,X' \stackrel{iid}{\sim} P$ and $Y,Y'\stackrel{iid}{\sim} Q$.
From the equality between \eqref{eq:mmd} and \eqref{eq:mmd2} we also
have 
\begin{equation}\label{eq:inner_data}
\langle \kkk_P, \kkk_Q \rangle_{\Hk} = \E \, \kk(X, Y).
\end{equation}
Thus, in practice, we can estimate the inner product between the 
embedded distributions 
by averaging the kernel function over sampled data.

The following important result shows that semimetrics of negative
type and symmetric positive semidefinite kernels are closely related
\cite{Berg1984}. Let $\rho: \mathcal{X} \times \mathcal{X} \to \mathbb{R}$,
and $x_0 \in \mathcal{X}$ an arbitrary but fixed point.
Define
\begin{equation}
\label{eq:kernel_semimetric}
\kk(x,y) = \tfrac{1}{2} \left[  \rho(x,x_0) + \rho(y,x_0) - \rho(x,y)\right].
\end{equation}
Thus, it can be shown that 
$\kk$ is positive definite if and only if $\rho$ is a semimetric
of negative type
\eqref{eq:negative_type}.
Here we have a family of kernels, one for each choice of $x_0$. Conversely,
if $\rho$ is a semimetric of negative type and $\kk$ is a kernel in this
family, then 
\begin{equation}
\label{eq:gen_kernel}
\begin{split}
\rho(x,y) &= \kk(x,x) + \kk(y,y) -2\kk(x,y) \\
&=  \| \kkk_x - \kkk_y \|^2_{\Hk},
\end{split}
\end{equation}
and the canonical feature map 
$\varphi: x \mapsto \kkk_x$ is injective \cite{Sejdinovic2013}.
When these conditions are satisfied, we say that the kernel $\kk$ 
generates the semimetric $\rho$. 
If two different kernels generate the same $\rho$ they are
equivalent kernels.

@@

Finally, let us recall the main formulas for test statistics
of equality of distributions \cite{Szkely2013}. 
Assume we have data $\mathbb{X} = \{ x_1,\dotsc, x_n \}$, where
$x_i \in \mathcal{X}$ and $\mathcal{X}$ is a space of negative type.
Consider a partition $\mathbb{X} = \bigcup_{j=1}^k \C_j$, with
$\C_i \cap \C_j = \emptyset$.
Each expectation in 
\eqref{eq:energy3}
can be computed 
through the function
\begin{equation}
\label{eq:g_def}
g (\C_i, \C_j) \equiv 
\dfrac{1}{n_i n_j}
\sum_{x \in \C_i} 
\sum_{y \in \C_j} \rho(x, y)
\end{equation}
where $n_i = |\C_i|$ is the number of elements in 
$\C_i$. 
The \emph{within energy dispersion} is defined by
\begin{equation}
\label{eq:within}
W \equiv
\sum_{j=1}^{k} \dfrac{n_j}{2} g(\C_j, \C_j),
\end{equation}
and the \emph{between-sample energy statistic} is defined by
\begin{equation}
\label{eq:between}
S \equiv
\sum_{1 \le  i < j \le k } \dfrac{n_i n_{j}}{2 n} \left[
2 g(\C_i, \C_j) - 
g(\C_i, \C_i) - 
g(\C_j, \C_j)
\right].
\end{equation}
Given a set of distributions
$\{ P_j\}_{j=1}^k$, where $x \in \C_j$ if and only if $x \sim P_j$, 
the quantity $S$ provides
a \emph{nonparametric} test statistic for equality of distributions
\cite{Szkely2013}.
When the sample size is large enough, $n\to \infty$,
under the null hypothesis $H_0: P_1=P_2=\dotsm=P_k$ we have that
$S\to 0$, 
and under
the alternative hypothesis $H_1: P_i \ne P_j$ for at least two $i\ne j$, 
we have that $S \to \infty$.
This test is nonparametric in the sense that it does not make any assumptions
about the distributions $P_j$.

One can make the analogy 
that every data point $ x \in \C_j$ form a massive body, 
whose total mass is characterized by the distribution function $P_j$. 
Then $S$ is a potential
energy of the from $S(P_1,\dotsc,P_k)$ which measures how different
are these
mass distributions, and achieves the ground state
$S=0$ when all bodies have the same mass distribution. The potential energy
$S$ increases as each body has different mass distribution than the other 
ones.

@@

This section contains the main results of this paper, where 
we formulate an optimization problem for clustering 
based on energy statistics and RKHS introduced in the previous section.

>>>>

\usepackage{algorithmic}
\usepackage{algorithm}
\usepackage[none]{hyphenat}

@@

\begin{abstract}
Energy statistics provides a nonparametric test for equality of distributions.
It was proposed by 
Sz\' ekely in the 80's
inspired by the Newtonian gravitational potential from classical mechanics. 
Energy statistics
was further generalized to probability 
distributions on arbitrary metric spaces,
and more recently a connection with reproducing kernel Hilbert spaces 
was established.
Nevertheless, although extensively used by the statistics community, it
has not been previously incorporated in machine learning problems.
In this paper, we consider the problem of clustering data from
an energy statistics theory perspective.
We provide a precise mathematical formulation yielding
a quadratically constrained quadratic program (QCQP).
We show the equivalence between the energy statistics clustering
formulation 
and the kernel $k$-means optimization problem (not to be confused with
kernel $k$-means algorithm). 
Therefore, our results imply a first principles derivation of 
kernel $k$-means problem
from energy statistics theory, thus bringing this important 
clustering method into a formal statistical framework.
Moreover, energy statistics is nonparametric, it usually fixes
the kernel choice,
and if prior information is available
it can be easily incorporated in the kernel construction.
We propose an iterative algorithm
to find local optimizers of the aforementioned QCQP based on the 
energy change
of moving points to different clusters. This algorithm 
is different but has the same computational cost as kernel $k$-means 
algorithm.
We compare it to
kernel $k$-means, standard $k$-means, and GMM/EM algorithms by 
providing carefully designed numerical experiments. The results
show that, in general, this algorithm outperforms
these most used clustering algorithms.
\end{abstract}

@@

Energy statistics is based on the energy distance between probability
distributions, which provides a notion of statistical potential energy
to statistical observations, in close analogy to Newton's gravitational
potential in classical mechanics. 
It provides a nonparametric test for equality of distribution,
such that minimum energy is achieved under equality of 
distributions. 
We refer the reader 
to \cite{Szkely2013}, and references therein, for an overview.
Energy statistics has been applied to several goodness-of-fit 
hypothesis tests, 
multi-sample tests of equality of distributions, analysis of variance
\cite{RizzoVariance}, and (nonlinear)
dependence tests through
distance covariance and distance correlation, which generalizes the Pearson
correlation coefficient. 
Moreover, energy statistics  was applied
to hierarchical clustering \cite{RizzoClustering} by extending Ward's
method of minimum variance.

More recently, distance covariance was 
generalized from Euclidean
spaces to metric spaces of strong negative type \cite{Lyons}.
Furthermore, a unifying framework establishing an equivalence
between generalized energy distances to maximum
mean discrepancies (MMD), which are distances between embeddings of
distributions in reproducing kernel Hilbert spaces (RKHS), was
established \cite{Sejdinovic2013}. This important work provides the missing
link
between techniques commonly used in the statistics literature, regarding
energy statistics, and techniques
commonly used in machine learning.

Given a dataset $\mathbb{X} = \{x_1,\dotsc,x_n \}$, where each
data point $x_i$ lives in a space $\mathcal{X}$, the clustering
problem consists in grouping these points into $k$ groups $\{
\C_1,\C_2,\dotsc,\C_k \}$, such that
points belonging to the same group are more ``similar'' to each other
than to points in  other groups. This intuitive notion already assumes
a ``metric'' able to measure the similarity between points. Clustering
is an unsupervised method, and one of the most important problems in machine
learning since it provides the first step towards automatically constructing
labels for previously unseen data. In practice
$k$-means is arguably the most used algorithm, and gaussian 
mixture models (GMM) through the expectation
maximization (EM) algorithm is also often used. Both are parametric, making 
strong assumptions about
the distribution of the data (isotropy and normality), and it assumes
that data points
lie in Euclidean space, $\mathcal{X} = \mathbb{R}^D$, hence
the similarity is based on Euclidean metric, 
$\| x_i - x_j \|^2 = (x_i - x_j)^\top(x_i-x_j) = 
\sum_{\ell=1}^D (x_{i,\ell} - x_{j,\ell})^2$. These algorithms provide
a good quality clustering when data is linearly separable in Euclidean space,
i.e. there is a clear resolution in the cluster centers
and points from different clusters do not overlap considerably.

To account for nonlinearly separable data, which may live
in an arbitrary non-Euclidean space, kernel methods are usually employed.
If a so-called Mercer kernel $\kk: \mathcal{X}\times \mathcal{X} \to \mathbb{R}$
is available, which is a symmetric and positive definite 
function, it guarantees that there exists a map $\varphi : \mathcal{X} \to
\mathcal{H}$ where $\mathcal{H}$ is a Hilbert space \cite{Mercer}. 
One can thus
compute similarities between data point through the inner product of
the feature space $\mathcal{H}$ and 
perform clustering in $\mathcal{H}$. The convenience of a Mercer
kernel is that the map $\varphi$ is only used implicitly, i.e. one
does not need to know $\varphi$, since the inner
of product of $\mathcal{H}$ can be computed only 
based on the kernel function $\kk(\cdot, \cdot)$ which acts on the original
space $\mathcal{X}$. 
This technique is known as the kernel trick.  
Kernel $k$-means problem 
is exactly $k$-means in the feature space
\cite{Smola,Girolami} (see also \cite{Filippone} for a survey of clustering
methods).
To cluster data that is not linearly
separable in $\mathcal{X}$,
kernel based methods exploit the nonlinear structure provided by
the kernel function, which implicitly maps data points to a usually higher
dimensional feature 
space, with the hope that they becomes separable by 
forming well-defined groups.
The choice of
kernel is crucial, and yet there is no principled method
for this. One must rely on ad hoc methods to find an appropriate kernel
which is data dependent.
As it stands, kernel $k$-means problem is an heuristic approach to clustering
in the sense that it is not derived from a statistical theory.

The main contribution of this paper is to provide a consistent formulation
to clustering based on energy statistics theory. 
The original goal is to provide a nonparametric
clustering method, since energy distance is nonparametric.
The basis of our work is the theory developed in 
\cite{Sejdinovic2013} which establishes an equivalence between energy
distance and RKHS. 
Based on this and on the multi-sample test
for equality of distributions, we show that
there is only one possible clustering optimization
problem consistent with the energy statistics formulation. 
We then show that this 
optimization problem is equivalent to a quadratically constrained quadratic
program (QCQP), which has the same trace maximization
form as kernel $k$-means,
spectral clustering, and some well-known graph partitioning problems
\cite{Dhillon2,Dhillon}. We show the equivalence between this approach
and kernel $k$-means optimization problem, implying that
the later can actually be derived from energy statistics, thus
bringing this important clustering method into a unified statistical theory.
However, it is important to note that 
energy statistics fixes a kernel to begin with.
From an algorithmic perspective, our contribution is to 
provide an iterative algorithm, which is different but has the same
complexity as kernel $k$-means algorithm. Our numerical results
provide compelling evidence that this algorithm 
is more accurate and robust than kernel $k$-means algorithm.
Moreover, these experiments 
puts in evidence the nonparametric aspect of energy
statistics based clustering, since it is able to perform accurately
on data having very different distributions, contrary to $k$-means and
GMM for instance.

Our work is organized as follows. In section~\ref{sec:background} we review
the necessary background on energy statistics and RKHS.
Section~\ref{sec:clustering_theory} contains the main theoretical 
results of this paper,
where we consider a clustering theory based on energy statistics, leading
to a QCQP which is NP-hard.
In section~\ref{sec:twoclass} we consider a simple example in one dimension,
where we propose an algorithm which requires no initialization.
In section~\ref{sec:algo} we briefly review kernel $k$-means algorithm,
and propose a new iterative algorithm to solve this QCQP.
Section~\ref{sec:numerics} contains some carefully designed numerical
experiments indicating that this algorithm outperforms kernel
$k$-means, standard $k$-means, and GMM/EM algorithms.
Our final conclusions are presented in section~\ref{sec:conclusion}.

@@

In this section we briefly review the main concepts from energy
statistics and its relation to reproducing kernel Hilbert spaces 
(RKHS) which form the basis of our work.
For more details we refer the reader
to \cite{Szkely2013} and \cite{Sejdinovic2013}.

Consider random variables in $\mathbb{R}^D$ 
such that $X,X' \stackrel{iid}{\sim} P$ and 
$Y,Y' \stackrel{iid}{\sim} Q$, where $P$ and $Q$ are cumulative
distribution functions with finite first moments. 
The quantity \cite{Szkely2013}
\begin{equation}
\label{eq:energy}
\Energy(P, Q) \equiv 2 \E \| X - Y\| - \E \| X - X' \| - \E \| Y - Y' \|,
\end{equation}
called \emph{energy distance}, 
is rotational invariant and nonnegative, $\Energy(P,Q) \ge 0$, where
equality
to zero holds if and only if $P = Q$.
Above, $\| \cdot \|$ denotes the
Euclidean norm in $\mathbb{R}^D$. 
Energy distance
provides a characterization of equality of distributions, and
$\Energy^{1/2}$ is
a metric on the space of distributions.

@@

It is important to 
mention that \eqref{eq:energy2} can be even further generalized.
Let $X, Y \in \mathcal{X}$,  where $\mathcal{X}$ is an arbitrary space endowed
with a \emph{semimetric of negative type}
$\rho: \mathcal{X}\times\mathcal{X} \to \mathbb{R}$, which is required
to satisfy
\begin{equation}
\label{eq:negative_type}
\sum_{i,j=1}^n c_i c_j \rho(X_i, X_j) \le 0,
\end{equation}
where $X_i \in \mathcal{X}$, and $c_i \in \mathbb{R}$ such that
$\sum_{i=1}^n c_i = 0$. Then, $\mathcal{X}$ is called a \emph{space of
negative type}.
We can thus replace $\mathbb{R}^D \to \mathcal{X}$ and 
$\| X - Y \| \to \rho(X , Y)$ in the definition \eqref{eq:energy}, obtaining
the generalized energy distance
\begin{equation}
\label{eq:energy3}
\Energy(P, Q) \equiv 2 \E \rho(X,Y) - \E \rho(X, X') - \E \rho(Y,Y').
\end{equation}
For spaces of negative type, there exists a Hilbert space $\mathcal{H}$ and
a map $\varphi: \mathcal{X} \to
\mathcal{H}$ such that
$\rho(X, Y) = \| \varphi(X) - \varphi(Y) \|_{\mathcal{H}}^2$, which
allows us to compute quantities related to probability distributions over
$\mathcal{X}$ in the Hilbert space $\mathcal{H}$.
Even though the semimetric 
$\rho$ may not satisfy the triangle inequality, 
$\rho^{1/2}$ does since it can be shown to be a legit metric.

There is an equivalence between energy distance, 
commonly used in statistics,
and distances between embeddings of distributions in 
RKHS, commonly used in machine learning. 
This equivalence was established
in \cite{Sejdinovic2013}. Let us first recall the definition of
RKHS. Let $\HH$ be a Hilbert space of real-valued functions
over $\mathcal{X}$. A function 
$\kk : \mathcal{X} \times \mathcal{X} \to 
\mathbb{R}$ is a reproducing kernel of $\HH$ if it satisfies
the following two conditions:
\begin{enumerate}
\item $\kkk_x \equiv \kk(\cdot, x) \in \HH$ 
for all $x \in \mathcal{X}$.
\item $\langle \kkk_x, f \rangle_{\HH} = f(x)$ for
all $x\in\mathcal{X}$ and $f\in \HH$.
\end{enumerate}
In other words, for any $x \in \mathcal{X}$ and any function $f \in \HH$,
there is a unique 
$\kkk_x \in \HH$ that reproduces $f(x)$ through the inner product
of $\HH$.
If such a \emph{kernel} 
function $\kk$ exists, then $\HH$ is called a RKHS. The above two 
properties immediately imply that $\kk$ is symmetric and positive
definite. Indeed, notice that
$\langle \kkk_x, \kkk_y \rangle = \kkk_y(x) = \kk(x,y)$, and since
this inner product is real,
$\langle \kkk_x, \kkk_y \rangle^* = \langle \kkk_y, \kkk_x \rangle = 
\langle \kkk_x, \kkk_y \rangle$, we immediately have that
the kernel is symmetric,
$\kk(y,x) = \kk(y,x)$. Moreover, for any $w \in
\HH$ we can write $w = \sum_{i=1}^n c_i \kkk_{x_i}$, where
$\{ \kkk_{x_i} \}_{i=1}^n$ is a basis of $\HH$. It follows that
$\langle w, w \rangle_{\HH}  = \sum_{i,j=1}^n c_i c_j \kk(x_i,x_j) \ge 0$,
showing that the kernel is positive definite. If $G$ is a matrix with
elements $G_{ij} = \kk(x_i,x_j)$, this is equivalent to $G$ being
positive semidefinite; $v^\top G \, v \ge 0$ for any vector
$v \in \mathbb{R}^n$.

The Moore-Aronszajn theorem 
\cite{Aronszajn}
establishes the converse of the above paragraph. 
For every symmetric
and positive definite function $\kk: \mathcal{X}\times \mathcal{X} \to
\mathbb{R}$, there is an associated RKHS $\Hk$ 
with reproducing
kernel $\kk$. The map $\varphi: x \mapsto \kkk_x \in \Hk$ is called
the canonical \emph{feature map}. Given a kernel $\kk$,
this theorem enables us to define an embedding of a probability measure
$P$ into the RKHS as follows: $P \mapsto \kkk_P \in
\Hk$ such that 
$\int f(x) d P(x) = \langle f, \kkk_P \rangle$ for all $f \in \Hk$,
or alternatively, $\kkk_P \equiv \int \kk( \, \cdot \,, x)  d P(x)$. 
We can now  introduce the 
notion of distance between two probability measures using the inner product
of $\Hk$. This is called the maximum mean discrepancy (MMD) and
is given by
\begin{equation}\label{eq:mmd}
\gamma_\kk(P,Q) \equiv \| \kkk_P - \kkk_Q \|_{\Hk},
\end{equation}
which can also be written as \cite{Gretton2012}
\begin{equation}\label{eq:mmd2}
\gamma_\kk^2(P,Q) = \E \kk(X,X') + \E \kk(Y,Y') - 2 \E \kk(X, Y)
\end{equation}
where $X,X' \stackrel{iid}{\sim} P$ and $Y,Y'\stackrel{iid}{\sim} Q$.
From the equality between \eqref{eq:mmd} and \eqref{eq:mmd2} we also
have 
\begin{equation}\label{eq:inner_data}
\langle \kkk_P, \kkk_Q \rangle_{\Hk} = \E \, \kk(X, Y).
\end{equation}
Thus, in practice, we can estimate the inner product between  
embedded distributions 
by averaging the kernel function over sampled data.

The following important result shows that semimetrics of negative
type and symmetric positive semidefinite kernels are closely related
\cite{Berg1984}. Let $\rho: \mathcal{X} \times \mathcal{X} \to \mathbb{R}$,
and $x_0 \in \mathcal{X}$ an arbitrary but fixed point.
Define
\begin{equation}
\label{eq:kernel_semimetric}
\kk(x,y) \equiv 
\tfrac{1}{2} \left[  \rho(x,x_0) + \rho(y,x_0) - \rho(x,y)\right].
\end{equation}
Then, it can be shown that 
$\kk$ is positive definite if and only if $\rho$ is a semimetric
of negative type
\eqref{eq:negative_type}.
Here we have a family of kernels, one for each choice of $x_0$. Conversely,
if $\rho$ is a semimetric of negative type and $\kk$ is a kernel in this
family, then 
\begin{equation}
\label{eq:gen_kernel}
\begin{split}
\rho(x,y) &= \kk(x,x) + \kk(y,y) -2\kk(x,y) \\
&=  \| \kkk_x - \kkk_y \|^2_{\Hk},
\end{split}
\end{equation}
and the canonical feature map 
$\varphi: x \mapsto \kkk_x$ is injective \cite{Sejdinovic2013}.
When these conditions are satisfied, we say that the kernel $\kk$ 
generates the semimetric $\rho$. 
If two different kernels generate the same $\rho$ they are
equivalent kernels.

@@

Finally, let us recall the main formulas for test statistic
of equality of distributions \cite{Szkely2013}. 
Assume we have data $\mathbb{X} = \{ x_1,\dotsc, x_n \}$, where
$x_i \in \mathcal{X}$, and $\mathcal{X}$ is a space of negative type.
Consider a partition $\mathbb{X} = \bigcup_{j=1}^k \C_j$, with
$\C_i \cap \C_j = \emptyset$.
Each expectation in 
\eqref{eq:energy3}
can be computed 
through the function
\begin{equation}
\label{eq:g_def}
g (\C_i, \C_j) \equiv 
\dfrac{1}{n_i n_j}
\sum_{x \in \C_i} 
\sum_{y \in \C_j} \rho(x, y)
\end{equation}
where $n_i = |\C_i|$ is the number of elements in 
$\C_i$. 
The \emph{within energy dispersion} is defined by
\begin{equation}
\label{eq:within}
W \equiv
\sum_{j=1}^{k} \dfrac{n_j}{2} g(\C_j, \C_j),
\end{equation}
and the \emph{between-sample energy statistic} is defined by
\begin{equation}
\label{eq:between}
S \equiv
\sum_{1 \le  i < j \le k } \dfrac{n_i n_{j}}{2 n} \left[
2 g(\C_i, \C_j) - 
g(\C_i, \C_i) - 
g(\C_j, \C_j)
\right],
\end{equation}
where $n = \sum_{j=1}^k n_j$.
Given a set of distributions
$\{ P_j\}_{j=1}^k$, where $x \in \C_j$ if and only if $x \sim P_j$, 
the quantity $S$ provides
a \emph{nonparametric test statistic} for equality of distributions
\cite{Szkely2013}.
When the sample size is large enough, $n\to \infty$,
under the null hypothesis $H_0: P_1=P_2=\dotsm=P_k$ we have that
$S\to 0$, 
and under
the alternative hypothesis $H_1: P_i \ne P_j$ for at least two $i\ne j$, 
we have that $S \to \infty$.
This test is nonparametric in the sense that it does not make any assumptions
about the distributions $P_j$.

One can make the analogy 
that every data point $ x \in \C_j$ form a massive body, 
whose total mass is characterized by the distribution function $P_j$.
The quantity $S$ is thus a potential
energy of the from $S(P_1,\dotsc,P_k)$ which measures how different
these
mass distributions are, and achieves the ground state
$S=0$ when all bodies have the same mass distribution. The potential energy
$S$ increases as bodies have different mass distributions.

@@

This section contains the main theoretical results of this paper, where 
we formulate an optimization problem for clustering 
based on energy statistics and RKHS introduced in the previous section.
